Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-13 cordillera-cycle.tex
Update to new Copernicus class.
Remove unit command where useless.

% Copernicus manuscript
\documentclass[tc, ms]{copernicus}

@@

% Copernicus discussion paper
%\documentclass[tcd, hvmath]{copernicus_discussions}

@@

% Coloured hyperlinks
\usepackage[colorlinks]{hyperref}

% Encoding
\usepackage[T1]{fontenc}

@@

% ----------------------------------------------------------------------
\begin{document}\hack{\sloppy}
% ----------------------------------------------------------------------

@@

% Authors
\author[1,2]{J.~Seguinot}
\author[3]{M.~Margold}
\author[2]{I.~Rogozhina}
\author[1]{A.~Stroeven}
\runningauthor{J.~Seguinot et~al.}
\correspondence{J.~Seguinot (julien.seguinot@natgeo.su.se)}

@@

% For Copernicus
\received{}
\accepted{}
\published{}

@@

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during
  marine isotope stages~4 (61.9--55.4\,\unit{kyr}) and~2
  (29.5--16.9\,\unit{kyr}). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional ocean sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,\unit{kyr}).

@@

For all proxy records, we firstly run a low-resolution simulation, using
a~10\,\unit{km}, 300 by 150~points horizontal grid for SSA and SIA velocities,
up to 51~enthalpy layers in the ice, and 31~temperature layers in the bedrock.
For the GRIP and EPICA records, we then run a second simulation using a higher
horizontal resolution of 6\,\unit{km}, up to 101~ice layers, and 61~bedrock
layers. In the rest of this paper, we refer to different model runs by the
proxy record used for palaeo-temperature forcing.

@@

Despite large differences in input temperature offset time-series, the model
output presents consistent features that can be observed across the range of
forcing data used. In all simulations, total ice volume is relatively low
during most of the glacial cycle, while two major glaciation events occur
during Marine Isotope Stage (MIS)~4 (61.9--55.4\,\unit{kyr}) and MIS~2
(29.5--16.9\,\unit{kyr}, Fig.~\ref{fig:timeseries}). A local ice minimum is
consistently attained during MIS~3 (52.2--45.6\,\unit{kyr}). However, the
magnitude and timing of these events depend significantly on the choice of
proxy record used for palaeo-temperature forcing.

@@

While simulations driven by the Antarctic (EPICA, Vostok) and GRIP records
reach a last glacial maximum in ice volume between 19.5 and 16.9\,\unit{kyr},
those steered by oceanic (ODP~1012, ODP~1020) and the NGRIP record attain it
several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
deglaciation prior to 20\,\unit{kyr} due to a local temperature maximum that is
inconsistent with all other records used. The ODP~1020 simulation predicts an
early last glacial maximum in ice volume at 29.5\,\unit{kyr}, followed by
slower deglaciation than modelled using other records. Finally, model runs
informed by Antarctic ice core records produce a continuous, rapid deglaciation
after the last glacial maximum, whereas the simulations driven by the GRIP
record produce a rapid deglaciation in three steps, including stand-stills and
readvances of the ice margin.

Despite the different timing, snapshots of model output show relatively
consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
different scaling factors applied to palaeo-temperature proxy records, the
ice sheet geometry at the last glacial maximum appears very similar from one
simulation to the next. It includes a central divide located at about
3\,500\,\unit{m} elevation along the spine of the Rocky Mountains. In all
simulations, a central ice cap persists over the Skeena mountains during MIS~3
between the two glaciation events. However, the size of this ice cap, as well
as the magnitude of the MIS~4 glaciation, depends sensitively on the choice of
palaeo-temperature proxy record used to drive the model.

@@

Independently of the choice of palaeo-temperature record used as forcing
time-series for the ice sheet model, our simulations consistently produce two
glaciation maxima during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2
(29.5--16.9\,\unit{kyr}). These events broadly correspond in timing to the
Gladstone (MIS~4) and McConell (MIS~2) glaciations documented by geological
evidence in the northern part of the model domain \aref, and to the Fraser
(MIS~2) glaciation in the southern part \aref.

@@

During most of the glacial cycle, modelled ice cover is restricted to disjoint
ice caps centred on major mountain ranges of the North American Cordillera
(Fig.~\ref{fig:duration}). In both simulations driven by the GRIP and EPICA
records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
Mountains exists for about 29\,\unit{kyr} in total over MIS~2 and~4.

@@

In our simulations, the timing of the last glacial maximum, and in turn that of
the deglaciation, depends on the palaeo-temperature time-series used to force
the model (Figs.~\ref{fig:timeseries} and ~\ref{fig:deglacseries}). The
6\,\unit{km}-resolution simulation forced by the EPICA record reaches a maximal
ice volume at 17.3\,\unit{kyr}, followed by a continuous deglaciation between
about 17 and 11\,\unit{kyr}. That driven by the GRIP record reaches maximal
volume at 19.1\,\unit{kyr}, and produces a deglaciation in three steps,
including the first stand-still between 16.6 and 14.5\,\unit{kyr}, and a late
glacial re-advance between 12.6 and 11.6\,\unit{kyr}.

@@

Although the predicted timing of glacier retreat differs, its patterns are
relatively consistent between the two simulations (Fig.~\ref{fig:deglac}). In
both runs, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated by 12\,\unit{kyr}, whereas significant
ice cover remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. At this stage, a late glacial re-advance
occurs in the GRIP simulation. Both simulations then show an unzipping of the
Liard Lowland, followed by ice retreat towards the Selwyn Mountains and the
Skeena Mountains, which host the last remains of the palaeo-ice sheet.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glaciation maxima during
MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}), two periods
corresponding to documented extensive palaeo-glaciations. This result is
independent of the choice of palaeo-temperature record used to approximate the
past evolution of climate forcing, and
can be seen as a first-order agreement between the model and geomorphological
evidence. However, the timing of the two glaciation peaks depends sensitively
on which record
is used to drive the model. The timing of the last glacial maximum is best
reproduced by the Antarctic record, and occurs too early in all simulations
that are driven by other records. The mismatch is greatest when using
oceanic records from the Pacific Northwest, which are affected by the
weakening of the California current during the last glacial maximum.

@@

Concerning the deglaciation phase, when most geological evidence is available,
none of the palaeo-temperature records used produces an ideal agreement between
model results and geological evidence. The timing of the last glacial maximum
and early deglaciation is generally best reproduced by forcing the model with
Antarctic records. However, the Younger Dryas re-advance is only reproduced
using temperature forcing from the GRIP record at places where it has been
documented. Nonetheless, the patterns of deglaciation are consistent between
both simulations driven by the GRIP and EPICA record. They show a rapid
deglaciation of the southern half of the ice
sheet, followed by unzipping of the Liard Lowland, and concentric retreat of
the ice margin towards the last palaeo-ice cover in the Skeena Mountains
during the early Holocene (10.9--9.5\,\unit{kyr}).

@@

% References
%\input{references.bbl}
\newpage

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters used to
           prepare temperature offset time-series used to force the ice sheet
           model through the last 120\,\unit{kyr}. $T_{[32;22]}$ refers to the
           mean temperature anomaly during the period -32 to~-22~\unit{kyr} after
           scaling.}
  \label{tab:records}
  {\begin{tabular}{lcccc}
    \tophline
    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Source\\
    \middlehline
    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C & \aref \\
    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C & \aref \\
    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

@@

% fig:timeseries
\begin{figure}
  \includegraphics{timeseries}
  \caption{Temperature offset time-series from ice core and sediment core
           records (Table~\ref{tab:records}) used as palaeo-climate forcing for
           the ice sheet model \textbf{(top)}, and modelled ice volume
           through the last 120\,\unit{kyr}, expressed in meters of sea-level
           equivalent \textbf{(bottom)}. Gray spans indicate Marine Isotope
           Stages (MIS) according to a global compilation of benthic
           \chem{\delta^{18}O} records \citep{lisiecki-raymo-2005}. Hatched
           rectangles highlight modelled ice volume extrema corresponding to
           MIS~4 (61.9--55.4\,\unit{kyr}), MIS~3 (52.2--45.6\,\unit{kyr}), and
           MIS~2 (last glacial maximum, 29.5--16.9\,\unit{kyr}). Dotted lines
           correspond to the GRIP and EPICA 6\,\unit{km}-resolution runs.
           \todo{add grid lines to improve readability}}
  \label{fig:timeseries}
\end{figure}

% fig:snapshots
\begin{figure}
  \includegraphics{snapshots}
  \caption{Snapshots of modelled surface topography (500\,\unit{m} contours)
           corresponding to the ice volume extrema indicated on
           Fig.~\ref{fig:timeseries}. Note the occurence of spatial similarities
           despite large differences in timing.}
  \label{fig:snapshots}
\end{figure}

% fig:duration
\begin{figure}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,\unit{kyr}.
           Note the irregular colour scale. A contiguous ice cover spanning
           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
           Rocky Mountains (RM) exists for about 29\,\unit{kyr} in both
           simulations. A central
           ice cover persists over the Skeena Mountains (SM) during most of the
           simulation. On the other hand, the maximal extent of the ice sheet
           generally corresponds to relatively short durations of ice cover.}
  \label{fig:duration}
\end{figure}

% fig:warmbase
\begin{figure}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,\unit{kyr}. Long ice cover durations combined with basal
           temperatures at the pressure-melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.
           \note{indicate location of Skeena Mountains.}}
  \label{fig:warmbase}
\end{figure}

% fig:erosion
\begin{figure}
  \includegraphics{erosion}
  \caption{Modelled cumulative basal displacement (integrand of basal velocity)
           over the last 120\,\unit{kyr}.
           \note{I am still unsure about the use of this figure. I may
                 remove it from this paper and keep the idea for later use.}}
  \label{fig:erosion}
\end{figure}

@@

% fig:deglac
\begin{figure}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote the Younger Dryas re-advance, which is more
           pronounced in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure}

% fig:deglacshots
\begin{figure}
  \includegraphics{deglacshots}
  \caption{Snapshots of modelled surface topography (200\,\unit{m} contours)
           and surface velocity (colour mapping) from the GRIP simulation,
           corresponding to the last glacial maximum (-19.5 kyr) and the last
           deglaciation.
           \note{should we have a similar figure using EPICA?}}
  \label{fig:deglacshots}
\end{figure}

% fig:profiles
\begin{figure}
  \includegraphics{profiles}
  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
           during deglaciation (22.0--8.0\,\unit{kyr}) in the GRIP 6\,\unit{km}
           simulation, corresponding to the four transects indicated in
           Fig.~\ref{fig:deglac}.
           \note{should we have a similar figure using EPICA?}}
  \label{fig:profiles}
\end{figure}

% fig:lastflow
\begin{figure}
  \includegraphics{lastflow}
  \caption{Modelled directions of the last basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout the simulation.
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.}
  \label{fig:lastflow}
\end{figure}

>>>>

% Copernicus manuscript
\documentclass[tc, manuscript]{copernicus}

@@

% Copernicus discussion paper
%\documentclass[tcd, hvmath, online]{copernicus}

@@

% Coloured hyperlinks
\hypersetup{colorlinks}

@@

% ----------------------------------------------------------------------
\begin{document}
\linenumbers
% ----------------------------------------------------------------------

@@

% Authors
\Author[1,2]{J.}{Seguinot}
\Author[3]{M.}{Margold}
\Author[2]{I.}{Rogozhina}
\Author[1]{A.~P.}{Stroeven}
\runningauthor{J.~Seguinot et~al.}
\correspondence{J.~Seguinot (julien.seguinot@natgeo.su.se)}

@@

% For Copernicus
\received{}
\pubdiscuss{}
\revised{}
\accepted{}
\published{}

@@

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving atmospheric
  forcing, all simulations produce two major glaciation events during
  marine isotope stages~4 (61.9--55.4\,kyr) and~2
  (29.5--16.9\,kyr). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional ocean sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remains of Cordilleran ice during the
  early Holocene (10.9--9.5\,kyr).

@@

For all proxy records, we firstly run a low-resolution simulation, using
a~10\,km, 300 by 150~points horizontal grid for SSA and SIA velocities,
up to 51~enthalpy layers in the ice, and 31~temperature layers in the bedrock.
For the GRIP and EPICA records, we then run a second simulation using a higher
horizontal resolution of 6\,km, up to 101~ice layers, and 61~bedrock
layers. In the rest of this paper, we refer to different model runs by the
proxy record used for palaeo-temperature forcing.

@@

Despite large differences in input temperature offset time-series, the model
output presents consistent features that can be observed across the range of
forcing data used. In all simulations, total ice volume is relatively low
during most of the glacial cycle, while two major glaciation events occur
during Marine Isotope Stage (MIS)~4 (61.9--55.4\,kyr) and MIS~2
(29.5--16.9\,kyr, Fig.~\ref{fig:timeseries}). A local ice minimum is
consistently attained during MIS~3 (52.2--45.6\,kyr). However, the
magnitude and timing of these events depend significantly on the choice of
proxy record used for palaeo-temperature forcing.

@@

While simulations driven by the Antarctic (EPICA, Vostok) and GRIP records
reach a last glacial maximum in ice volume between 19.5 and 16.9\,kyr,
those steered by oceanic (ODP~1012, ODP~1020) and the NGRIP record attain it
several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
deglaciation prior to 20\,kyr due to a local temperature maximum that is
inconsistent with all other records used. The ODP~1020 simulation predicts an
early last glacial maximum in ice volume at 29.5\,kyr, followed by
slower deglaciation than modelled using other records. Finally, model runs
informed by Antarctic ice core records produce a continuous, rapid deglaciation
after the last glacial maximum, whereas the simulations driven by the GRIP
record produce a rapid deglaciation in three steps, including stand-stills and
readvances of the ice margin.

Despite the different timing, snapshots of model output show relatively
consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
different scaling factors applied to palaeo-temperature proxy records, the
ice sheet geometry at the last glacial maximum appears very similar from one
simulation to the next. It includes a central divide located at about
3\,500\,m elevation along the spine of the Rocky Mountains. In all
simulations, a central ice cap persists over the Skeena mountains during MIS~3
between the two glaciation events. However, the size of this ice cap, as well
as the magnitude of the MIS~4 glaciation, depends sensitively on the choice of
palaeo-temperature proxy record used to drive the model.

@@

Independently of the choice of palaeo-temperature record used as forcing
time-series for the ice sheet model, our simulations consistently produce two
glaciation maxima during MIS~4 (61.9--55.4\,kyr) and MIS~2
(29.5--16.9\,kyr). These events broadly correspond in timing to the
Gladstone (MIS~4) and McConell (MIS~2) glaciations documented by geological
evidence in the northern part of the model domain \aref, and to the Fraser
(MIS~2) glaciation in the southern part \aref.

@@

During most of the glacial cycle, modelled ice cover is restricted to disjoint
ice caps centred on major mountain ranges of the North American Cordillera
(Fig.~\ref{fig:duration}). In both simulations driven by the GRIP and EPICA
records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
Mountains exists for about 29\,kyr in total over MIS~2 and~4.

@@

In our simulations, the timing of the last glacial maximum, and in turn that of
the deglaciation, depends on the palaeo-temperature time-series used to force
the model (Figs.~\ref{fig:timeseries} and ~\ref{fig:deglacseries}). The
6\,km-resolution simulation forced by the EPICA record reaches a maximal
ice volume at 17.3\,kyr, followed by a continuous deglaciation between
about 17 and 11\,kyr. That driven by the GRIP record reaches maximal
volume at 19.1\,kyr, and produces a deglaciation in three steps,
including the first stand-still between 16.6 and 14.5\,kyr, and a late
glacial re-advance between 12.6 and 11.6\,kyr.

@@

Although the predicted timing of glacier retreat differs, its patterns are
relatively consistent between the two simulations (Fig.~\ref{fig:deglac}). In
both runs, the southern half of the modelling domain, including the Puget
Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
Columbia, are completely deglaciated by 12\,kyr, whereas significant
ice cover remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
in the northern half of the domain. At this stage, a late glacial re-advance
occurs in the GRIP simulation. Both simulations then show an unzipping of the
Liard Lowland, followed by ice retreat towards the Selwyn Mountains and the
Skeena Mountains, which host the last remains of the palaeo-ice sheet.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glaciation maxima during
MIS~4 (61.9--55.4\,kyr) and MIS~2 (29.5--16.9\,kyr), two periods
corresponding to documented extensive palaeo-glaciations. This result is
independent of the choice of palaeo-temperature record used to approximate the
past evolution of climate forcing, and
can be seen as a first-order agreement between the model and geomorphological
evidence. However, the timing of the two glaciation peaks depends sensitively
on which record
is used to drive the model. The timing of the last glacial maximum is best
reproduced by the Antarctic record, and occurs too early in all simulations
that are driven by other records. The mismatch is greatest when using
oceanic records from the Pacific Northwest, which are affected by the
weakening of the California current during the last glacial maximum.

@@

Concerning the deglaciation phase, when most geological evidence is available,
none of the palaeo-temperature records used produces an ideal agreement between
model results and geological evidence. The timing of the last glacial maximum
and early deglaciation is generally best reproduced by forcing the model with
Antarctic records. However, the Younger Dryas re-advance is only reproduced
using temperature forcing from the GRIP record at places where it has been
documented. Nonetheless, the patterns of deglaciation are consistent between
both simulations driven by the GRIP and EPICA record. They show a rapid
deglaciation of the southern half of the ice
sheet, followed by unzipping of the Liard Lowland, and concentric retreat of
the ice margin towards the last palaeo-ice cover in the Skeena Mountains
during the early Holocene (10.9--9.5\,kyr).

@@

% References
%\bibliographystyle{copernicus}
%\bibliography{references.bib}
\newpage

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters used to
           prepare temperature offset time-series used to force the ice sheet
           model through the last 120\,kyr. $T_{[32;22]}$ refers to the
           mean temperature anomaly during the period -32 to~-22~kyr after
           scaling.}
  \label{tab:records}
  {\begin{tabular}{lcccc}
    \tophline
    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Source\\
    \middlehline
    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C & \aref \\
    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C & \aref \\
    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

@@

% fig:timeseries
\begin{figure*}
  \includegraphics{timeseries}
  \caption{Temperature offset time-series from ice core and sediment core
           records (Table~\ref{tab:records}) used as palaeo-climate forcing for
           the ice sheet model \textbf{(top)}, and modelled ice volume
           through the last 120\,kyr, expressed in meters of sea-level
           equivalent \textbf{(bottom)}. Gray spans indicate Marine Isotope
           Stages (MIS) according to a global compilation of benthic
           \chem{\delta^{18}O} records \citep{lisiecki-raymo-2005}. Hatched
           rectangles highlight modelled ice volume extrema corresponding to
           MIS~4 (61.9--55.4\,kyr), MIS~3 (52.2--45.6\,kyr), and
           MIS~2 (last glacial maximum, 29.5--16.9\,kyr). Dotted lines
           correspond to the GRIP and EPICA 6\,km-resolution runs.
           \todo{add grid lines to improve readability}}
  \label{fig:timeseries}
\end{figure*}

% fig:snapshots
\begin{figure*}
  \includegraphics{snapshots}
  \caption{Snapshots of modelled surface topography (500\,m contours)
           corresponding to the ice volume extrema indicated on
           Fig.~\ref{fig:timeseries}. Note the occurence of spatial similarities
           despite large differences in timing.}
  \label{fig:snapshots}
\end{figure*}

% fig:duration
\begin{figure*}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,kyr.
           Note the irregular colour scale. A contiguous ice cover spanning
           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
           Rocky Mountains (RM) exists for about 29\,kyr in both
           simulations. A central
           ice cover persists over the Skeena Mountains (SM) during most of the
           simulation. On the other hand, the maximal extent of the ice sheet
           generally corresponds to relatively short durations of ice cover.}
  \label{fig:duration}
\end{figure*}

% fig:warmbase
\begin{figure*}
  \includegraphics{warmbase}
  \caption{Modelled duration of warm-based ice cover during the last
           120\,kyr. Long ice cover durations combined with basal
           temperatures at the pressure-melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.
           \note{indicate location of Skeena Mountains.}}
  \label{fig:warmbase}
\end{figure*}

% fig:erosion
\begin{figure*}
  \includegraphics{erosion}
  \caption{Modelled cumulative basal displacement (integrand of basal velocity)
           over the last 120\,kyr.
           \note{I am still unsure about the use of this figure. I may
                 remove it from this paper and keep the idea for later use.}}
  \label{fig:erosion}
\end{figure*}

@@

% fig:deglac
\begin{figure*}
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote the Younger Dryas re-advance, which is more
           pronounced in the GRIP-driven simulation.}
  \label{fig:deglac}
\end{figure*}

% fig:deglacshots
\begin{figure*}
  \includegraphics{deglacshots}
  \caption{Snapshots of modelled surface topography (200\,m contours)
           and surface velocity (colour mapping) from the GRIP simulation,
           corresponding to the last glacial maximum (-19.5 kyr) and the last
           deglaciation.
           \note{should we have a similar figure using EPICA?}}
  \label{fig:deglacshots}
\end{figure*}

% fig:profiles
\begin{figure}
  \includegraphics{profiles}
  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
           during deglaciation (22.0--8.0\,kyr) in the GRIP 6\,km
           simulation, corresponding to the four transects indicated in
           Fig.~\ref{fig:deglac}.
           \note{should we have a similar figure using EPICA?}}
  \label{fig:profiles}
\end{figure}

% fig:lastflow
\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled directions of the last basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout the simulation.
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.}
  \label{fig:lastflow}
\end{figure*}
